### 2024-05-16 journal/response.txt

*** should add new comments to GC code similar to new comments in basic code

@@

No, this was a typo.

> lines 552 - 567, right column, and lines 686 - 711, right column.
> Adding explicit references to the lines of the code on Fig. 5 - 7
> would help to follow the explanation.

Done.

> line 558, right column. The sentence "the block's element field is
> $e$ to indicate it represents Enqueue(e) operation..." is
> confusing. It's not immediately clear why element $e$ indicates
> that the operation is Enqueue. Probably formulate it "the block's
> element field $e\neq null$ indicates..."

This paragraph has been rewritten.

> lines 594 - 608 (lines 24 - 37 in the code). Add a more detailed
> comments that explain the Refresh operation.

> line 619 (line 44 in the code). It's not clear why the size of the
> queue is calculated correctly. Why if some of the dequeues that are
> counted by num_deq are empty and do not change the size of the
> queue ? Some intuition here would be helpful.

Some explanation has been added in the text (line 526) [*** check line #].

> line 620 (line 46 in the code). Suggestion: consider to put the
> check whether the new block B contains no operations
> (num_deqs+num_enqs==0) immediately after the calculation of num_enq
> and num_deqs. First, complete the calculations that depend on the
> values of num_enq and num_deq, and then advance to the check
> whether v is root. If the change will be done, change the order of
> the explanation in lines 715 - 723, right column, accordingly.

### 2024-05-16 journal/algorithm.tex
handling some more comments from reviewer 1

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute 
\begin{eqnarray*}
B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
the number of enqueues and dequeues in~$B$
\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
The enqueues of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
Then the dequeues of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements
(unless the size becomes 0 at some point during these dequeues).  Thus, we have}
\begin{eqnarray*}
B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
			&& \hspace*{19mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).
\end{eqnarray*}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
    \State \Call{Append}{\var{B}}\label{enqAppend}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
    \State \Call{Append}{\var{B}}\label{deqAppend}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$}\label{invokeIndexDequeue}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}\label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh}
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
    \If{$\var{B} = \nl$} \Return{\tr} \label{addOP} 
	\Else
	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}} \label{cas}
    	\State \Call{Advance}{\var{v, h}}\label{advance}
    	\State \Return{ \var{result}}
	\EndIf
\EndFunction{Refresh}
\end{algorithmic}
\caption{Queue implementation's main routines.\label{code1}}
\end{figure*}

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{39}
\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \Comment{create new block $B$ for \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields\label{initNewBlock}
    	\eleft\ \assign\ $\var{v}.\fld{left}.\head - 1$,\label{createEndLeft}
    	\eright\ \assign\ $\var{v}.\fld{right}.\head - 1$,\label{createEndRight}\\
		\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{enq}}\label{createSumEnq},\\
		\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{deq}} + \var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
    \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmpty} 
    \Else\ \Return \var{B}
    \EndIf
\EndFunction{CreateBlock}

@@

An \opemph{Enqueue}(\var{e}) 
\rev{creates a new Block at line \ref{enqNew} and sets its fields to indicate
that the Block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
field.
The call to \op{Append}, described below, on line \ref{enqAppend}
appends the \block\ to the process's leaf and ensures it is propagated to the root.}

A \opemph{Dequeue} also 
\rev{creates a new Block at line \ref{deqNew} and sets its fields to indicate
that the Block represents a \op{Dequeue}.
It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
it is propagated to the root.
Then,} the \op{Dequeue} calls \op{IndexDequeue} to compute
its position in the root 
 and then calls \op{FindResponse} to compute its response.

@@

\opemph{CreateBlock}(\var{v, i}) is used
by \op{Refresh} to construct a new block $B$ to be installed in \var{v}.\fld{blocks}[\var{i}].
The \eleft\ and \eright\ fields store the indices of the last blocks appended to \var{v}'s
children, obtained by reading the \fld{head} index in \var{v}'s children.
Since the \fld{sum\sub{enq}} field should store the number of enqueues in
\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{B}.\eright].\fld{sum\sub{enq}}$.
Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
subtracting  the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
Then, if \var{B}
is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
represents the number of elements in the queue after the operations in the block are performed.
Finally, if the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
 there is no need to install~it.

>>>>

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute 
\begin{eqnarray*}
B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
the number of enqueues and dequeues in~$B$
\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
The enqueues of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
Then the dequeues of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
(If subtracting this many elements would make the new size negative because there are more dequeues than there are elements, then the new size is 0.)  Thus, we have}
\begin{eqnarray*}
B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
			&& \hspace*{19mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).
\end{eqnarray*}

@@

\Function{void}{Enqueue}{\typ{Object} \var{e}} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
		\fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{enqAppend}
\EndFunction{Enqueue}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{Root}.\fld{blocks}}}\label{invokeIndexDequeue}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \op{Dequeue} should return}}\label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  \rev{if first fails, try again}}
        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \Comment{\rev{help advance $v$'s children's \var{head} indices, if necessary}}\label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}}\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlock}
    \If{$\var{B} = \nl$} \Return{\tr}\Comment{\rev{if no new operations to propagate to $v$, do nothing}}\label{addOP} 
	\Else
	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}}\Comment{\rev{try to add $B$ to \var{v}.\fld{blocks}}}\label{cas}
    	\State \Call{Advance}{\var{v, h}} \Comment{\rev{advance $v.head$ from $h$ to $h+1$}}\label{advance}
    	\State \Return{ \var{result}}\Comment{\rev{return true iff \op{CAS} succeeded}}
	\EndIf
\EndFunction{Refresh}
\end{algorithmic}
\caption{Queue implementation's main routines.\label{code1}}
\end{figure*}

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{39}
\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \Comment{create new block $B$ for \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields\label{initNewBlock}
    	\eleft\ \assign\ $\var{v}.\fld{left}.\head - 1$,\label{createEndLeft}
    	\eright\ \assign\ $\var{v}.\fld{right}.\head - 1$,\label{createEndRight}\\
		\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{enq}}\label{createSumEnq},\\
		\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{deq}} + \var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
    \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmpty} 
    \ElsIf{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
    \EndIf
    \State \Return \var{B}
\EndFunction{CreateBlock}

@@

An \opemph{Enqueue}(\var{e}) 
\rev{creates a new Block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
that the Block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
field.
The call to \op{Append}, described below, on line \ref{enqAppend}
appends the \block\ to the process's leaf and ensures it is propagated to the root.}

A \opemph{Dequeue} also 
\rev{creates a new Block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
that the Block represents a \op{Dequeue}.
It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
it is propagated to the root.
Then,} the \op{Dequeue} calls \op{IndexDequeue} to compute
its position in the root 
 and then calls \op{FindResponse} to compute its response.

@@

\opemph{CreateBlock}(\var{v, i}) is used
by \op{Refresh} to construct a new block $B$ to be installed in \var{v}.\fld{blocks}[\var{i}].
The \eleft\ and \eright\ fields store the indices of the last blocks appended to \var{v}'s
children, obtained by reading the \fld{head} index in \var{v}'s children.
Since the \fld{sum\sub{enq}} field should store the number of enqueues in
\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{B}.\eright].\fld{sum\sub{enq}}$.
Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
subtracting  the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
\rev{If the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
 there is no need to install~it.
Otherwise,} if \var{B}
is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
represents the number of elements in the queue after the operations in the block are performed.
